### 2017-07-01 motivation.tex
RC figs edited.
edits to motivation

We have developed and implemented our ideas in an OCaml embedded DSL
that supports a \C{DB} monad to define and compose database
computations given in terms of SQL operations; the monadic \C{do}
syntax is borrowed from Haskell.  Besides the usual \C{bind} and
\C{return} combinators, monad offers an \C{atomically} combinator that
executes a database computation as an atomic transaction and returns
the result.

@@

Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
\C{new\_order} transaction.  TPC-C is a well-known Online Transaction
Processing (OLTP) benchmark that models an order-processing system for
a wholesale parts supply business. The business logic is captured in 5
database transactions that operate on 9 tables. \C{new\_order} is one
such transaction that operates on \C{District}, \C{Order},
\C{New\_order}, \C{Stock}, and \C{Order\_line} tables. The transaction
acts on the behalf of a customer, whose id is \C{c\_id}, to place a
new order for the given set of items (\C{item\_reqs}), to be served by
a warehouse under the district identified by \C{d\_id}. The
transaction does so by invoking appropriate SQL functionality,
captured by various calls to functions under the \C{SQL} module. All
\C{SQL} functions take the table name (a nullary constructor) as their
first argument. The higher-order \C{SQL.select1} function accepts a
boolean function that describes the selection criteria, and returns
any record that meets the criteria (it models the SQL query \C{SELECT
  \ldots\xspace LIMIT 1}). \C{SQL.update} also accepts a Boolean
function (3rd argument) to select the records to be updated. Its
2$^{nd}$ argument is a function that maps each selected record to a new
(updated) record. \C{SQL.insert} inserts a given record into the
specified table in the database.

@@

TPC-C defines multiple invariants, called \emph{consistency
  conditions}, over the state of the application in the database. One
such consistency condition is the requirement that for a given order
\C{o}, the \emph{order-line-count} field (\C{o.o\_ol\_cnt}) should
reflect the number of order lines under the order, i.e., the number of
\C{Order\_line} records whose \C{ol\_o\_id} is equal to \C{o.o\_id}.
In a sequential execution, it is easy to see how this condition is
preserved.  A new \C{Order} record is added with its \C{o\_id}
distinct from the existing order ids, and its \C{o\_ol\_cnt} is set to
be equal to the size of \C{item\_reqs} set. The \C{foreach} loop runs
once for each \C{item\_req}, adding a new \C{Order\_line} record for
each requested item, with its \C{ol\_o\_id} field set to
\C{o\_id}. Thus, at the end of the loop, the number of \C{Order\_line}
records in the database, whose \C{ol\_o\_id} is equal to \C{o\_id}, is
equal to the size of the \C{item\_req} set, which in turn is equal to
the \C{Order} record's \C{o\_ol\_cnt} field, thus preserving the
consistency condition.

Because the aforementioned reasoning is reasonably simple to perform
manually, verfiying that soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions.  Under weak isolation\footnote{Weak isolation
  doesn't violate atomicity as long as the witnessed effects are those
  of committed transactions}, however, interferences of various kinds
are permitted.  Although the verification problem for weakly isolated
transactions would appear to be superficially similar to the
verification of (racy) concurrent programs~\cite{concurrentGC...},
weak isolation introduces new challenges arising from the use of
transactions, and new opportunities arising from the fact that the
store abstraction used by transaction is a relational database, not
low-level memory.  Our proof framework addresses these challenges and
exploits these opportunities to facilitiate fully automated
correctness proofs.

To illustrate, consider the behavior of the \C{new\_order} transaction
when executed with a \emph{Read Committed} (RC) isolation level, the
default isolation level in PostgreSQL, a widely used open-source
database system.  An RC transaction is isolated from \emph{dirty
  writes}, i.e., writes of uncommitted transactions, but is allowed to
witness the writes of concurrent transactions as soon as they are
committed. Thus, with two concurrent instances of the \C{new\_order}
transaction (call them $T_1$ and $T_2$), both concurrently placing new
orders for different customers under the same district (\C{d\_id}),
PostgreSQL admits the two executions shown in
Fig.~\ref{fig:new_order_execs}.

\begin{figure}[!h]
\centering
\subcaptionbox {
  {\sc rc} Execution 1
  \label{fig:motiv-eg-1-b}
} [
  0.55\columnwidth
] {
  \includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
}
%\hspace*{0.5in}
\subcaptionbox {
  {\sc rc} Execution 2
  \label{fig:motiv-eg-1-a}
}{
  \includegraphics[scale=0.45]{Figures/motiv-eg-1-a}
}
  \caption{\small RC executions involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction of
  Fig.~\ref{fig:new_order_code}. 
  Each instance reads the \C{d\_id} \C{District} record twice, second
  time to (atomically) update the \C{d\_next\_o\_id} field.}
\label{fig:rc-ex}
\end{figure}

The figure depicts an execution as a series of read, write, and commit
operations. In the execution on the left, the \C{new\_order} instance
$T_1$ (green) reads the \C{d\_next\_o\_id} field of the district
record for \C{d\_id}, but before it increments the field, another
\C{new\_order} instance ($T_2$) begins its execution and commits. Note
that $T_2$ reads the same \C{d\_next\_o\_id} value as $T_1$, and
inserts new \C{Order} and \C{Order\_line} records with their \C{o\_id}
and \C{ol\_o\_id} fields (resp.) equal to \C{d\_next\_o\_id}. $T_2$
also increments the \C{d\_next\_o\_id} field, which $T_1$ has already
acccessed. This is allowed because reads do not obtain a mutually
exclusive lock on most databases, including PostgreSQL. After $T_2$'s
commit, $T_1$ resumes execution and adds new \C{Order} and
\C{Order\_line} fields with the same order id as $T_1$. Thus, by the
end of the execution, \C{Order\_line} records inserted by $T_1$ and
$T_2$ all bear the same order id. There are also two \C{Order} records
with the same district id (\C{d\_id}) and order id, none of whose
\C{o\_ol\_cnt} reflects the actual number of \C{Order\_line} records
inserted with that order id.  This clearly violates TPC-C's consistency
condition.

@@

While reasoning in terms of anomalies is cumbersome, and as the above
example shows often inadequate, reasoning about weak isolation in
terms of low-level traces~\cite{adyaphd,gotsmanconcur15} confounds
high-level reasoning and automation.  Another alternative would
interleave weak isolation implementation details within the program,
yielding a (more-or-less) conventional concurrent program that can be
then subject to classical concurrent verification methods.
Considering the size and complexity of real-world transaction systems,
this approach is unlikely to scale.

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a unified framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
$\C{new_order}$ is sound when executed under a \emph{Repeatable Read}
(RR) isolation level.  PostgreSQL executes an RR transaction by taking
a (conceptual) snapshot of the database state before the transaction
begins that is used by reads performing the transaction.  When the
transaction attempts to update a record, a check is made to determine
if the current version of the record in the database is same as the
snapshot version. If it is, an exclusive lock is obtained on the
record, and the update is performed. If the record is already locked,
the transaction blocks until the lock is released.  When the lock is
obtained, is held exclusively until the transaction commits or rolls
back. If the aforementioned version check fails (i.e., the database
has a later version than the snapshot), the transaction is rolled back
and re-executed.

\begin{figure}[t]
\includegraphics[scale=0.5]{Figures/RR-postgres}
\caption{Database state transitions corresponding to an execution of
  an RR transaction $T_1$ on Postgres. $T_2$ and $T_3$ are concurrent
  transactions. State has three data items, $a$, $b$ and $c$. }
\label{fig:rr-postgres}
\end{figure}
\begin{figure}[t]
\includegraphics[scale=0.5]{Figures/RR-abstract}
  \caption{An abstract execution equivalent to the one in
  Fig.~\ref{fig:rr-postgres}. Database state remains invariant as long as
  $T_1$ executes. The effects of concurrent transactions ($T_2$ and
  $T_3$) are witnessed just before the commit.}
\label{fig:rr-abstract}
\end{figure}

Although this implementation, comprising many thousands of lines code,
is highly involved, its semantic behavior in terms of how it effects
transitions on the database state is fairly simple.  Since a
transaction's reads are always served from a snapshot, no state
changes are witnessed while the execution is in progress. Thus,
insofar as an RR transaction is concerned, the database state does not
change during the execution.  Uncommitted writes are recorded in a
transaction-local state.  When the transaction commits, the local
state is atomically written to the global state to yield a global
state that reflects the transaction's updates.  However, unlike a
strongly isolated serializable transaction, the commit operation is
performed against the current state of the database, not the
snapshot. Thus, after the transaction finishes execution, but before
it commits, the transaction is able to witness the effects of all
concurrent transactions.  The PostgreSQL RR implementation effectively
constrains this transition.

We can axiomatize this operational description by observing that (a)
due to the version check, the current transaction cannot update a data
item that was already updated by a concurrent transaction, and (b) due
to the use of exclusive write locks, a data item updated by the
current transaction cannot be overwritten by a concurrent transaction.
If $\Delta$ is the state of the database when an RR transactions
finishes, and $\Delta_c$ is the state visible to the transaction at
the point of commit, we know the transition from $\Delta$ to
$\Delta_c$ (written $\Delta \longrightarrow \Delta_c$) cannot exhibit
effects from any concurrent transactions that write to the same data
items as the current RR transaction.  Similarly, if $\delta$ denotes
a local log that relates transaction variables being written with their updated values,
then $\forall x\in\mathit{dom}(\delta)$, $\Delta_c(x) =
\Delta(x)$. To summarize, the operational semantics of PostgresSQL's RR
implementation can be captured as an axiomatization over transitions of
the database state ($\Delta \longrightarrow \Delta'$) during the
lifetime an RR transaction ($T$):
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a transaction
reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these updates. An update
statement filters the tuples that match the search criteria from $\Delta$
and computes the updated tuples that are to be
added to the local database. Thus, its state transformer (call it
$T_U$) is the following function on sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $\Delta$$\bind$(fun s -> if table(s) = stock && s.s_i_id = item_req.ol_i_id 
                                 then {{s with s_qty = k1}} (* a singleton set *)
                                 else $\emptyset$)
\end{ocaml}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
Observe that
$T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
F_U(\Delta)$. The transformer ($T_I(\Delta,\delta)$)
for the subsequent \C{insert} statement can be similarly calculated to
be of the form $\delta \cup F_I(\Delta)$.
Their composition gives the state transformer of the loop body:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $F_U(\Delta) \,\cup\, F_I(\Delta)$
\end{ocaml}
Let $F_{body}(\Delta) = F_U(\Delta) \,\cup\, F_I(\Delta)$.   The
transformer for the \C{foreach} loop can now be computed as following:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ item_reqs$\bind$(fun item_req -> $F_{body}(\Delta)$)
\end{ocaml}
Observe that the transformer captures the precise semantics of the
loop as a formula in the domain of sets extended with a bind
function. The advantage in doing so is that we can now make use of a
semantics-preserving translation from this domain to first-order
logic, allowing us to leverage SMT solvers for automatic proofs
without having to infer loop invariant, or provide intermediate
assertions.  Sec.~\ref{sec:automation} describes this translation. In
the exposition thus far, we assumed $\Delta$ remains invaraint, which
is clearly not the case when we admit concurrency.  Necessary
concurrency extensions of the state transformer semantics is also
covered in Sec.~\ref{sec:automation}.  The immediate two sections,
however, focus on laying theoretical foundations for reasoning about
weak isolation.

>>>>

We have developed and implemented our ideas in an embedded DSL in
OCaml that supports a \C{DB} monad to define and compose database
computations given in terms of SQL operations.  Besides the usual
\C{bind} and \C{return} combinators, the monad offers an
\C{atomically} combinator that executes a database computation as an
atomic transaction and returns the result.

@@

Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
\C{new\_order} transaction written in this language.  TPC-C is a
well-known Online Transaction Processing (OLTP) benchmark that models
an order-processing system for a wholesale parts supply business. The
business logic is captured in 5 database transactions that operate on
9 tables; \C{new\_order} is one such transaction that operates on
\C{District}, \C{Order}, \C{New\_order}, \C{Stock}, and
\C{Order\_line} tables. The transaction acts on the behalf of a
customer, whose id is \C{c\_id}, to place a new order for the given
set of items (\C{item\_reqs}), to be served by a warehouse under the
district identified by \C{d\_id}.  Fig.~\ef{fig:scheme} illustrates
the relationship among these different tables.

The transaction manages order placement by invoking appropriate SQL
functionality, captured by various calls to functions defined by the
\C{SQL} module. All \C{SQL} functions take the table name (a nullary
constructor) as their first argument. The higher-order \C{SQL.select1}
function accepts a boolean function that describes the selection
criteria, and returns any record that meets the criteria (it models
the SQL query \C{SELECT \ldots\xspace LIMIT 1}). \C{SQL.update} also
accepts a Boolean function (its 3$^{rd}$ argument) to select the records to be
updated. Its 2$^{nd}$ argument is a function that maps each selected
record to a new (updated) record. \C{SQL.insert} inserts a given
record into the specified table in the database.

%%%SJ: Reviewers may not understand what primary and foreign keys are,
%%%or why they are important.  There are also some fields (e.g., ol_i_id)
%%%that are not described either in the caption or the text.

@@

TPC-C defines multiple invariants, called \emph{consistency
  conditions}, over the state of the application in the database. One
such consistency condition is the requirement that for a given order
\C{o}, the \emph{order-line-count} field (\C{o.o\_ol\_cnt}) should
reflect the number of order lines under the order; this is the number
of \C{Order\_line} records whose \C{ol\_o\_id} field is the same as
\C{o.o\_id}.  In a sequential execution, it is easy to see how this
condition is preserved.  A new \C{Order} record is added with its
\C{o\_id} distinct from existing order ids, and its \C{o\_ol\_cnt} is
set to be equal to the size of the \C{item\_reqs} set. The \C{foreach}
loop runs once for each \C{item\_req}, adding a new \C{Order\_line}
record for each requested item, with its \C{ol\_o\_id} field set to
\C{o\_id}. Thus, at the end of the loop, the number of \C{Order\_line}
records in the database, whose \C{ol\_o\_id} is equal to \C{o\_id} is
equal to the size of the \C{item\_req} set, which in turn is equal to
the \C{Order} record's \C{o\_ol\_cnt} field, thus preserving the
consistency condition.

Because the aforementioned reasoning is reasonably simple to perform
manually, verfiying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes.
Under weak isolation\footnote{Weak isolation doesn't violate atomicity
  as long as the witnessed effects are those of committed
  transactions}, however, interferences of various kinds are
permitted.  Although the verification problem for weakly isolated
transactions would appear to be superficially similar to the
verification of (racy) concurrent programs~\cite{concurrentGC...},
weak isolation introduces new challenges arising from the use of
transactions, and new opportunities arising from the fact that the
store abstraction used by transactions is a relational database, not
low-level memory.

To illustrate some of these challenges, consider the behavior of the
\C{new\_order} transaction when executed with a \emph{Read Committed}
(RC) isolation level, the default isolation level in PostgreSQL, a
widely used open-source database system.  An RC transaction is
isolated from \emph{dirty writes}, i.e., writes of uncommitted
transactions, but is allowed to witness the writes of concurrent
transactions as soon as they are committed. Thus, with two concurrent
instances of the \C{new\_order} transaction (call them $T_1$ and
$T_2$), both concurrently placing new orders for different customers
under the same district (\C{d\_id}), PostgreSQL admits the two
executions shown in Fig.~\ref{fig:new_order_execs}.

\begin{figure}[!h]
\centering
\subcaptionbox {
  {\sc rc} Execution 1
  \label{fig:motiv-eg-1-b}
} [
  0.55\columnwidth
] {
  \includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
}
%\hspace*{0.5in}
\subcaptionbox {
  {\sc rc} Execution 2
  \label{fig:motiv-eg-1-a}
}{
  \includegraphics[scale=0.45]{Figures/motiv-eg-1-a}
}
  \caption{\small RC executions involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Each instance reads the \C{d\_id} \C{District} record twice, the second
  time to (atomically) update the \C{d\_next\_o\_id} field.}
\label{fig:new_order_execs}
\end{figure}

The figure depicts an execution as a series of transactional read, write, and commit
operations. In the execution on the left, the \C{new\_order} instance
$T_1$ (green) reads the \C{d\_next\_o\_id} field of the district
record for \C{d\_id}, but before it increments the field, another
\C{new\_order} instance ($T_2$) begins its execution and commits. Note
that $T_2$ reads the same \C{d\_next\_o\_id} value as $T_1$, and
inserts new \C{Order} and \C{Order\_line} records with their \C{o\_id}
and \C{ol\_o\_id} fields (resp.) equal to \C{d\_next\_o\_id}. $T_2$
also increments the \C{d\_next\_o\_id} field, which $T_1$ has already
acccessed. This is allowed because reads do not obtain a mutually
exclusive lock on most databases, including PostgreSQL. After $T_2$'s
commit, $T_1$ resumes execution and adds new \C{Order} and
\C{Order\_line} fields with the same order id as $T_1$. Thus, by the
end of the execution, \C{Order\_line} records inserted by $T_1$ and
$T_2$ all bear the same order id. There are also two \C{Order} records
with the same district id (\C{d\_id}) and order id, none of whose
\C{o\_ol\_cnt} reflects the actual number of \C{Order\_line} records
inserted with that order id.  This clearly violates TPC-C's consistency
condition.

@@

While reasoning in terms of anomalies is cumbersome, and as the above
example shows often inadequate, reasoning about weak isolation in
terms of low-level traces~\cite{adyaphd,gotsmanconcur15} complicates
high-level reasoning.  A possible alternative would interleave weak
isolation implementation details within the program, yielding a
(more-or-less) conventional concurrent program that can be then
subject to classical concurrent verification methods.  Considering the
size and complexity of real-world transaction systems, this strategy
is unlikely to scale.

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under a \emph{Repeatable Read}
(RR) isolation level.  PostgreSQL executes an RR transaction by taking
a (conceptual) snapshot of the database state before the transaction
begins that is used by reads performing the transaction.  When the
transaction attempts to update a record, a check is made to determine
if the current version of the record in the database is the same as the
snapshot version. If it is, an exclusive lock is obtained on the
record, and the update is performed. If the record is already locked,
the transaction blocks until the lock is released.  When the lock is
obtained, is held exclusively until the transaction commits or rolls
back. If the aforementioned version check fails (i.e., the database
has a later version than the snapshot), the transaction is rolled back
and re-executed.

\begin{figure}[t]
\includegraphics[scale=0.5]{Figures/RR-postgres}
\caption{Database state transitions corresponding to an execution of
  an RR transaction $T_1$ on PostgresSQL. $T_2$ and $T_3$ are concurrent
  transactions. The database state has three data items, $a$, $b$ and $c$. }
\label{fig:rr-postgres}
\end{figure}
\begin{figure}[t]
\includegraphics[scale=0.5]{Figures/RR-abstract}
  \caption{An abstract execution equivalent to the one in
  Fig.~\ref{fig:rr-postgres}. The database state remains invariant as long as
  $T_1$ executes. The effects of concurrent transactions ($T_2$ and
  $T_3$) are witnessed just before the commit.}
\label{fig:rr-abstract}
\end{figure}

Although this implementation, comprising many thousands of lines of
code, is highly involved, its semantic behavior in terms of how it
effects transitions on the database state is fairly simple.  Since a
transaction's reads are always served from a snapshot, no state
changes are witnessed while the execution is in progress. Thus,
insofar as an RR transaction is concerned, the database state does not
change during the execution.  Uncommitted writes are recorded in a
transaction-local state.  When the transaction commits, the local
state is atomically written to the global state to yield a global
state that reflects the transaction's updates.  However, unlike a
strongly isolated serializable transaction, the commit operation is
performed against the current state of the database, not the
snapshot. Thus, after the transaction finishes execution, but before
it commits, the transaction is able to witness the effects of all
concurrent transactions that committed before it.  The PostgreSQL RR
implementation effectively constrains this transition.

We can axiomatize this operational description by observing that (a)
due to the version check, the current transaction cannot update a data
item that was already updated by a concurrent transaction, and (b) due
to the use of exclusive write locks, a data item updated by the
current transaction cannot be overwritten by a concurrent transaction.
If $\Delta$ is the state of the database when an RR transaction
finishes, and $\Delta_c$ is the state visible to the transaction at
the point of commit, we know the transition from $\Delta$ to
$\Delta_c$ (written $\Delta \longrightarrow \Delta_c$) cannot exhibit
effects from any concurrent transactions that write to the same data
items as the current RR transaction.  Similarly, if $\delta$ denotes
a local log that relates transaction variables being written with their updated values,
then $\forall x\in\mathit{dom}(\delta)$, $\Delta_c(x) =
\Delta(x)$. To summarize, the operational semantics of PostgresSQL's RR
implementation can be captured as an axiomatization over transitions of
the database state ($\Delta \longrightarrow \Delta'$) during the
lifetime an RR transaction ($T$):
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a transaction
reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these updates. An update
statement filters the tuples that match the search criteria from $\Delta$
and computes the updated tuples that are to be
added to the local database. Thus, its state transformer (call it
$T_U$) is the following function on sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $\Delta$$\bind$(fun s -> if table(s) = stock && s.s_i_id = item_req.ol_i_id 
                                 then {{s with s_qty = k1}} (* a singleton set *)
                                 else $\emptyset$)
\end{ocaml}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
Observe that
$T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
F_U(\Delta)$. The transformer ($T_I(\Delta,\delta)$)
for the subsequent \C{insert} statement can be similarly calculated to
be of the form $\delta \cup F_I(\Delta)$.
Their composition gives the state transformer of the loop body:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $F_U(\Delta) \,\cup\, F_I(\Delta)$
\end{ocaml}
Let $F_{body}(\Delta) = F_U(\Delta) \,\cup\, F_I(\Delta)$.   The
transformer for the \C{foreach} loop can now be computed as following:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ item_reqs$\bind$(fun item_req -> $F_{body}(\Delta)$)
\end{ocaml}
Observe that the transformer captures the precise semantics of the
loop as a formula in the domain of sets extended with a bind
function. The advantage in doing so is that we can now make use of a
semantics-preserving translation from this domain to first-order
logic, allowing us to leverage SMT solvers for automatic proofs
without having to infer loop invariant, or provide intermediate
assertions.  Sec.~\ref{sec:automation} describes this translation. In
the exposition thus far, we assumed $\Delta$ remains invaraint, which
is clearly not the case when we admit concurrency.  Necessary
concurrency extensions of the state transformer semantics is also
covered in Sec.~\ref{sec:automation}.  The following two sections,
however, focus on laying theoretical foundations for reasoning about
weak isolation.
